
# Summary report for: `r spp_summ$spp_name`


## Overview of BBS data

Number of routes: `r spp_summ$n.routes`  

Number of outliers: `r spp_summ$n.outliers`

Number of buffered routes: `r spp_summ$n.buffer`

```{r route_map, echo=FALSE, fig.width=12, fig.height=10}
PlotRoutes(spp_summ$alpha)
```


```{r summary_table, results='asis', echo=FALSE}
# From the ms functions:
tab <- summary_table(data_study$data, baad_dictionary)
kable(tab[names(tab) != "Studies"], "markdown", align="l", padding=10)
```

```{r world_map, echo=FALSE, fig.width=12, fig.height=10}
drawWorldPlot(data_study$data)
```

And locally within the country:

```{r country_map, echo=FALSE, fig.width=12, fig.height=10}
map_sites_all_countries(data_study$data)
```

The sites sampled are:

```{r location_level_info, echo=FALSE, comment='', results='asis'}
kable(location_level_info(data_study$data), "markdown", align="l", padding=2)
```

The growing conditions of sampled plants was:

```{r stand_level_info, echo=FALSE, comment='', results="asis"}
kable(stand_level_info(data_study$data), "markdown", align="l", padding=2)
```

## Species sampled

```{r species_level_info, echo=FALSE, comment='', results="asis"}
kable(species_level_info(data_study$data), "markdown", align="l", padding=2)
```

## Methods used

```{r metadata, echo=FALSE, comment='', results="asis"}
metadata <- data_study$metadata
writeLines(sprintf("**%s**: %s\n", metadata$Topic, metadata$Description))
```

## Plots of data

This is how the study `r study_name` fits in the entire dataset (grey). each colour represents a species. A legend of species names with colours is included at the end for reports with 1 < n < 20 species.

```{r variable_plots, fig.height=12, fig.width=12, echo=FALSE}
report_variable_plots(data_study, baad_data, baad_dictionary)
```
